### 2023-04-09 main/chapter_theory.tex
add higgs and citations

The Standard Model Lagrangian can be decomposed into this sum of
(partial Lagrangian) terms:
\begin{equation}
\label{eqn:theory_sm_lagrangian_top_sum}
\mathcal{L}_\mathrm{SM} =
\mathcal{L}_\mathrm{YM}
+ \mathcal{L}_\mathrm{WD}
+ \mathcal{L}_\mathrm{Yu}
+ \mathcal{L}_\mathrm{H}
,
\end{equation}
which we define below;
$\mathcal{L}_\mathrm{YM}$ corresponds to the gauge bosons,
$\mathcal{L}_\mathrm{WD}$ corresponds to the fermions and their gauge
interactions,
$\mathcal{L}_\mathrm{Yu}$ corresponds to the Yukawa interactions between the
fermions and the Higgs,
and
$\mathcal{L}_\mathrm{H}$ corresponds to the Higgs boson and its potential.

\paragraph{Yang-Mills.}
The Yang-Mills summand $\mathcal{L}_\mathrm{YM}$ implements the gauge groups
$SU\!(3)_C$ (colour), $SU\!(2)_L$ (weak isospin), and $U\!(1)_Y$ (hypercharge),
each with a nested term respectively:
\begin{equation}
\label{eqn:theory_sm_lagrangian_top_ym}
\mathcal{L}_\mathrm{YM} =
\mathcal{L}_\mathrm{QCD}
+ \mathcal{L}_\mathrm{I_W}
+ \mathcal{L}_\mathrm{Y}
.
\end{equation}
All three Yang-Mills terms take similar forms:
\begin{align}
\label{eqn:theory_sm_lagrangian_top_ym_terms}
\mathcal{L}_\mathrm{QCD} &=
-\frac{1}{4g^2_3}
\sum_{A=1}^8 G^A_{\mu\nu} G^{\mu\nu A}
,\\
\mathcal{L}_\mathrm{I_W} &=
-\frac{1}{4g^2_2}
\sum_{a=1}^3 A^a_{\mu\nu} F^{\mu\nu a}
,\textrm{ and}\\
\mathcal{L}_\mathrm{Y} &=
-\frac{1}{4g^2_1} B{\mu\nu} B^{\mu\nu}
.
\end{align}
Here, $g_3$, $g_2$, and $g_3$ are dimensionless constants.
They relate to colour, weak isospin, and hypercharge, respectively.
Furthermore, for colour,
\begin{equation}
G^A_{\mu\nu} =
\partial_\mu A^A_\nu - \partial_\nu A^A_\mu - f^{ABC} A^B_\mu A^C_\nu
,
\end{equation}
with $A, B, C \in \{1, 2, ..., 8\}$ indexing the eight gluon fields $A^B_\mu$ and
the $SU\!(3)_C$ structure function $f^{ABC}$.
For weak isospin,
\begin{equation}
F^a_{\mu\nu} =
\partial_\mu W^a_\nu - \partial_\nu W^a_\mu - \epsilon^{abc} W^b_\mu W^c_\nu
,
\end{equation}
with $a, b, c \in \{1, 2, 3\}$ indexing the vector bosons $W^a_\mu$ and
the $SU\!(2)_L$ structure function $\epsilon^{abc}$.
For hypercharge with the vector boson $B$,
\begin{equation}
B_{\mu\nu} =
\partial_\mu B_\nu - \partial_\nu B_\mu
.
\end{equation}

\paragraph{Weyl-Dirac.}
The Weyl-Dirac summand $\mathcal{L}_\mathrm{WD}$ implements the fermions and
their gauge interactions, and can be written:
\begin{equation}
\begin{split}
\mathcal{L}_\mathrm{WD} =
\sum_i^3
\Bigl(
& L^\dagger_i \sigma^\mu \mathcal{D}_\mu L_i
+ \bar e^\dagger_i \sigma^\mu \mathcal{D}_\mu \bar e_i \\
& + \mathbf{Q}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{Q}_i
+ \mathbf{\bar u}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{\bar u}_i
+ \mathbf{\bar d}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{\bar d}_i
\Bigr)
,
\end{split}
\end{equation}
where the $\mathcal{D}_\mu$ expressions corresponding to the fermion fields:
$L_i$ for lepton doublets,
$\bar e_i$ for lepton singlets,
$\mathbf{Q}_i$ for quark doublets,
and $\mathbf{\bar u}_i$ and $\mathbf{\bar d}_i$ for up- and down-type quark
singlets, respectively.
The three fermion generations are indexed by $i$.
The quark fields come in colour triplets.

\paragraph{Yukawa.} \TODO{}

\paragraph{Higgs.} \TODO{}

\TODO{add citations}
%http://www.physics.ufl.edu/~ramond/JourneysChapter2_CUP.pdf

>>>>

The Standard Model Lagrangian can be decomposed into this sum of
(partial Lagrangian) terms:
\begin{equation}
\label{eqn:theory_sm_lagrangian_top_sum}
\mathcal{L}_\mathrm{SM} =
\mathcal{L}_\mathrm{YM}
+ \mathcal{L}_\mathrm{WD}
+ \mathcal{L}_\mathrm{Yukawa}
+ \mathcal{L}_\mathrm{Higgs}
,
\end{equation}
which we define below;
$\mathcal{L}_\mathrm{YM}$ corresponds to the gauge bosons,
$\mathcal{L}_\mathrm{WD}$ corresponds to the fermions and their gauge
interactions,
$\mathcal{L}_\mathrm{Yukawa}$ corresponds to the Yukawa interactions between the
fermions and the Higgs,
and
$\mathcal{L}_\mathrm{Higgs}$ corresponds to the Higgs boson and its
potential~\cite{ramond1999journeys,pdg2022ynf}.

\paragraph{Yang-Mills.}
The Yang-Mills summand $\mathcal{L}_\mathrm{YM}$ implements the gauge groups
$SU\!(3)_C$ (colour), $SU\!(2)_L$ (weak isospin), and $U\!(1)_Y$ (hypercharge),
each with a nested term respectively:
\begin{equation}
\label{eqn:theory_sm_lagrangian_top_ym}
\mathcal{L}_\mathrm{YM} =
\mathcal{L}_\mathrm{QCD}
+ \mathcal{L}_\mathrm{I_W}
+ \mathcal{L}_\mathrm{Y}
.
\end{equation}
All three Yang-Mills terms take similar forms:
\begin{align}
\label{eqn:theory_sm_lagrangian_top_ym_terms}
\mathcal{L}_\mathrm{QCD} &=
-\frac{1}{4g^2_3}
\sum_{A=1}^8 G^A_{\mu\nu} G^{\mu\nu A}
,\\
\mathcal{L}_\mathrm{I_W} &=
-\frac{1}{4g^2_2}
\sum_{a=1}^3 A^a_{\mu\nu} F^{\mu\nu a}
,\textrm{ and}\\
\mathcal{L}_\mathrm{Y} &=
-\frac{1}{4g^2_1} B_{\mu\nu} B^{\mu\nu}
.
\end{align}
Here, $g_3$, $g_2$, and $g_3$ are dimensionless constants.
They relate to colour, weak isospin, and hypercharge, respectively.
Furthermore, for colour,
\begin{equation}
G^A_{\mu\nu} =
\partial_\mu A^A_\nu - \partial_\nu A^A_\mu - f^{ABC} A^B_\mu A^C_\nu
,
\end{equation}
with $A, B, C \in \{1, 2, ..., 8\}$ indexing the eight gluon fields $A^B_\mu$ and
the $SU\!(3)_C$ structure function $f^{ABC}$.
For weak isospin,
\begin{equation}
F^a_{\mu\nu} =
\partial_\mu W^a_\nu - \partial_\nu W^a_\mu - \epsilon^{abc} W^b_\mu W^c_\nu
,
\end{equation}
with $a, b, c \in \{1, 2, 3\}$ indexing the vector bosons $W^a_\mu$ and
the $SU\!(2)_L$ structure function $\epsilon^{abc}$.
For hypercharge with the vector boson $B$,
\begin{equation}
B_{\mu\nu} =
\partial_\mu B_\nu - \partial_\nu B_\mu
.
\end{equation}

\paragraph{Weyl-Dirac.}
The Weyl-Dirac summand $\mathcal{L}_\mathrm{WD}$ implements the fermions and
their gauge interactions, and can be written:
\begin{equation}
\begin{split}
\mathcal{L}_\mathrm{WD} =
\sum_i^3
\Bigl(
& L^\dagger_i \sigma^\mu \mathcal{D}_\mu L_i
+ \bar e^\dagger_i \sigma^\mu \mathcal{D}_\mu \bar e_i \\
& + \mathbf{Q}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{Q}_i
+ \mathbf{\bar u}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{\bar u}_i
+ \mathbf{\bar d}^\dagger_i \sigma^\mu \mathcal{D}_\mu \mathbf{\bar d}_i
\Bigr)
,
\end{split}
\end{equation}
where the $\mathcal{D}_\mu$ expressions corresponding to the fermion fields:
$L_i$ for lepton doublets,
$\bar e_i$ for lepton singlets,
$\mathbf{Q}_i$ for quark doublets,
and $\mathbf{\bar u}_i$ and $\mathbf{\bar d}_i$ for up- and down-type quark
singlets, respectively.
The three fermion generations are indexed by $i$.
The quark fields are in colour triplets.

\paragraph{Yukawa.}
The Yukawa term implements masses for the charged leptons by coupling
them to the scalar Higgs field $H$:
\begin{equation}
\mathcal{L}_\mathrm{Yukawa} =
i\hat L_i \bar e_j H^* Y^{[e]}_{ij}
+ i\mathbf{\widehat Q}_i \mathbf{\bar d}_j H^* Y^{[d]}_{ij}
+ i\mathbf{\widehat Q}_i \mathbf{\bar u}_j \tau_2 H Y^{[u]}_{ij}
+ [\textrm{c.c.}]
,
\end{equation}
in which $\tau_2$ contains weak isospin indices,
and $Y^{[e]}_{ij}$, $Y^{[d]}_{ij}$, $Y^{[u]}_{ij}$ are Yukawa coupling matrices
with shape $3\times3$ for the three fermion generations.

\paragraph{Higgs.}
Finally, the Higgs term,
\begin{equation}
\mathcal{L}_\mathrm{Higgs} =
(\mathcal{D}_\mu H)^\dagger (\mathcal{D}^\mu H)
- \left[ -\mu^2 H^\dagger H + \lambda (H^\dagger H)^2 \right]
,
\end{equation}
contains the dynamics and potential for the scalar Higgs doublet.
The potential terms are isolated into square brackets to highlight the negative
sign on the mass term $\mu^2$, which with the quartic term contributes to
spontaneous symmetry breaking~\cite{rubakov1999classical}.
